---
title: "<span style='background-color: #3498db; color: white; padding: 10px; display: block; font-size: 28px; text-align: center; border-radius: 5px; box-shadow: 0 4px 6px rgba(0, 0, 0, 0.1);'>Exploratory Data Analysis of WMATA Transit System</span>"
author: "Tomato Smoothie🍅"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
    code_folding: hide
    css: |
      h1, h2, h3, h4 {
        color: #2c3e50;
        background-color: #f8f9fa;
        padding: 10px;
        border-left: 5px solid #3498db;
        font-weight: bold;
      }
      h1 {
        border-left: 5px solid #e74c3c;
      }
      h2 {
        border-left: 5px solid #3498db;
        margin-top: 30px;
      }
      h3 {
        border-left: 5px solid #2ecc71;
      }
---

```{r setup, include=FALSE}
# Configure global options for all code chunks
knitr::opts_chunk$set(
  echo = TRUE,          # Show code
  warning = FALSE,      # Hide warnings
  message = FALSE,      # Hide messages
  fig.width = 10,       # Default figure width
  fig.height = 6,       # Default figure height
  fig.align = "center"  # Center figures
)
```

## 📊 Introduction

This document presents a comprehensive exploratory data analysis (EDA) of the Washington Metropolitan Area Transit Authority (WMATA) system. The analysis examines multiple aspects of WMATA's ridership patterns including:

- Daily and monthly ridership trends
- Station-level usage patterns across the Metrorail system
- Peak vs. off-peak travel behavior
- Fare technology adoption (tap vs. non-tap entries)
- Travel patterns by time period

The insights derived from this analysis can help WMATA planners optimize service schedules, allocate resources efficiently, and improve the passenger experience across the network.

## 🔧 Data Preparation

### 📚 Load Required Libraries

```{r libraries}
# Data manipulation libraries
library(dplyr)      # For data transformation
library(tidyr)      # For data cleaning
library(readr)      # For reading CSV files

# Date and time handling
library(lubridate)  # For date manipulation

# Visualization libraries
library(ggplot2)    # For creating graphs
library(gridExtra)  # For arranging multiple plots
library(reshape2)   # For reshaping data
library(forcats)    # For factor manipulation
library(stringr)    # For string manipulation
```

### 📥 Import Data Sources

```{r load-data, message=FALSE, warning=FALSE}
# Import main ridership datasets
total_entries_exits <- read_csv("Total_Entries_and_Exits_Processed.csv")
entries_over_time <- read_csv("Entries_Over_Time_Processed.csv")
entries_by_tap <- read_csv("Entries_by_Tap_and_noTap_Processed.csv")

# Import aggregated boardings data
boardings_by_day <- read_csv("Boardings_by_Day_of_Week_Processed.csv")
boardings_by_month <- read_csv("Cleaned_Boardings_Data.csv")
boardings_weekends <- read_csv("Boardings_Weekends_Processed.csv")

# Import time-based and station-based data
boardings_time_period <- read_csv("Boardings_by_Time_Period_Processed.csv")
entries_by_year <- read_csv("Entries_by_Year_Processed.csv")
boardings_by_route <- read_csv("Boardings_by_Route_Table_Processed.csv")
```

### 🧹 Data Cleaning and Preprocessing

```{r data-cleaning}
# ---- Handle date and day columns ----
# Identify appropriate column names based on what's available in the data
date_col_name <- if("Date" %in% names(total_entries_exits)) "Date" else "Column_0"
day_col_name <- if("Day" %in% names(total_entries_exits)) "Day" else "Column_0"

# Extract and parse date information from text columns if needed
if("Column_0" %in% names(total_entries_exits) && !("Date" %in% names(total_entries_exits))) {
  # Check if Column_0 contains date strings (e.g., "Tuesday, March 11, 2025")
  sample_dates <- head(total_entries_exits$Column_0)
  date_pattern <- "\\w+,\\s+\\w+\\s+\\d+,\\s+\\d{4}"
  
  if(any(grepl(date_pattern, sample_dates))) {
    # Parse dates by removing the day of week prefix
    total_entries_exits$Date <- mdy(gsub("^\\w+,\\s+", "", total_entries_exits$Column_0))
    
    # Extract day of week from the full string
    total_entries_exits$Day <- sub(",.*$", "", total_entries_exits$Column_0)
  }
}

# ---- Clean numeric fields ----
# Convert string numbers with commas to numeric values
if(is.character(total_entries_exits$Entries)) {
  total_entries_exits$Entries <- as.numeric(stringr::str_replace_all(total_entries_exits$Entries, ",", ""))
}
if(is.character(total_entries_exits$Exits)) {
  total_entries_exits$Exits <- as.numeric(stringr::str_replace_all(total_entries_exits$Exits, ",", ""))
}

# Verify data quality after conversion
if(any(is.na(total_entries_exits$Entries)) || any(is.na(total_entries_exits$Exits))) {
  warning("NAs introduced during number conversion - check data format")
}

# ---- Standardize categorical variables ----
# Ensure days of week are properly ordered
day_levels <- c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")
if("Day" %in% names(total_entries_exits)) {
  total_entries_exits$Day <- factor(total_entries_exits$Day, levels = day_levels)
}

# ---- Standardize column names across datasets ----
# Process boardings_by_day dataframe
if("Day of Week" %in% names(boardings_by_day)) {
  boardings_by_day <- boardings_by_day %>% rename(Day = `Day of Week`)
}

# Identify and standardize value column in boardings_by_day
boardings_column_names <- names(boardings_by_day)
if(length(boardings_column_names) >= 2) {
  if(!"Boardings" %in% boardings_column_names) {
    boardings_value_col <- boardings_column_names[2]  # Assuming the value column is the second one
    boardings_by_day <- boardings_by_day %>% rename(Boardings = all_of(boardings_value_col))
  }
} else {
  warning("boardings_by_day doesn't have enough columns")
}

# Process station names in entries_by_tap
if("Station Name" %in% names(entries_by_tap)) {
  entries_by_tap <- entries_by_tap %>% rename(Station = `Station Name`)
}

# Create proper month ordering for time-based analysis
if("Month" %in% names(boardings_by_month)) {
  month_levels <- month.name  # Built-in R vector with month names
  boardings_by_month$Month <- factor(boardings_by_month$Month, levels = month_levels)
}
```

## 🔍 Exploratory Data Analysis

### 📋 Data Structure Overview

```{r check-data}
# Display structure of main datasets to understand their composition
str(total_entries_exits)
str(boardings_by_day)
```

### 📊 Summary Statistics

```{r summary-stats}
# Generate summary statistics for key datasets
summary(total_entries_exits)
summary(boardings_by_day)
summary(entries_by_year)
```

## 📈 Visualizing Temporal Trends

### 📆 Ridership Patterns Over Time

```{r time-trends, fig.cap="Daily entries and exits over time"}
# Check if Date column exists and has valid data
if("Date" %in% names(total_entries_exits) && 
   is.numeric(total_entries_exits$Entries) && 
   is.numeric(total_entries_exits$Exits)) {
  
  # Convert to Date type if it's not already
  if(!inherits(total_entries_exits$Date, "Date")) {
    total_entries_exits$Date <- as.Date(total_entries_exits$Date)
  }
  
  # Create time series plot of entries and exits
  ggplot(total_entries_exits, aes(x = Date)) +
    geom_line(aes(y = Entries, color = "Entries"), size = 1) +
    geom_line(aes(y = Exits, color = "Exits"), size = 1) +
    labs(
      title = "Entries and Exits Over Time",
      subtitle = "Daily ridership patterns",
      x = "Date", 
      y = "Count"
    ) +
    scale_color_manual(values = c("Entries" = "blue", "Exits" = "red")) +
    theme_minimal() +
    theme(
      legend.title = element_blank(),
      legend.position = "bottom",
      plot.title = element_text(face = "bold"),
      axis.title = element_text(face = "bold")
    )
} else {
  cat("Cannot create time trends plot - missing Date column or invalid data\n")
}
```

### 📅 Day of Week Patterns

```{r day-of-week, fig.cap="Ridership distribution by day of week"}
# Using the correct column name for boardings
if("Day" %in% names(boardings_by_day) && "Boardings" %in% names(boardings_by_day)) {
  ggplot(boardings_by_day, aes(x = Day, y = Boardings, fill = Day)) +
    geom_bar(stat = "identity") +
    labs(
      title = "Total Boardings by Day of the Week",
      subtitle = "Weekday vs. weekend ridership patterns", 
      x = "Day", 
      y = "Boardings"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold"),
      axis.title = element_text(face = "bold")
    )
} else {
  # Create a simplified version using original column names
  ggplot(boardings_by_day, aes(x = boardings_by_day[[1]], y = boardings_by_day[[2]])) +
    geom_bar(stat = "identity", fill = "steelblue") +
    labs(
      title = "Total Boardings by Day of the Week", 
      x = names(boardings_by_day)[1], 
      y = names(boardings_by_day)[2]
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold"),
      axis.title = element_text(face = "bold")
    )
}
```

### 📅 Monthly Ridership Patterns

```{r monthly-patterns, fig.cap="Monthly ridership distribution"}
boardings_data <- read_csv("Cleaned_Boardings_Data.csv")

# Aggregate data by month (sum of avg_daily across years and tap types)
boardings_by_month <- boardings_data %>%
  group_by(month) %>%
  summarize(Value = sum(avg_daily, na.rm = TRUE)) %>%
  rename(Month = month)

# Ensure proper month ordering
month_order <- c("January", "February", "March", "April", "May", "June", 
                "July", "August", "September", "October", "November", "December")
boardings_by_month$Month <- factor(boardings_by_month$Month, levels = month_order)

# Create the plot
ggplot(boardings_by_month, aes(x = Month, y = Value, fill = Month)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Boardings by Month", 
    subtitle = "Seasonal patterns in transit usage",
    x = "Month", 
    y = "Average Daily Boardings"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold"),
    legend.position = "none"
  )
```

## 🚉 Station-Level Analysis

### 🥇 Top Stations by Ridership

```{r top-stations, fig.cap="Top 10 stations by total boardings"}
# Check if the file has the expected structure based on our inspection
if("Station Name" %in% names(boardings_by_route)) {
  # Create a new column with total boardings across all time periods
  boardings_by_route <- boardings_by_route %>%
    mutate(Total = `AM Peak (Open-9:30am)` + 
           `Midday (9:30am-3pm)` + 
           `PM Peak (3pm-7pm)` + 
           `Evening (7pm-12am)` + 
           `Late Night (12am-Close)`)
  
  # Get top stations
  top_stations <- boardings_by_route %>%
    arrange(desc(Total)) %>%
    slice_head(n = 10)
  
  # Create the plot
  ggplot(top_stations, aes(x = reorder(`Station Name`, -Total), y = Total, fill = `Station Name`)) +
    geom_bar(stat = "identity") +
    labs(
      title = "Top 10 Stations by Total Boardings", 
      subtitle = "Major activity centers in the transit network",
      x = "Station", 
      y = "Boardings"
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "none",
      plot.title = element_text(face = "bold"),
      axis.title = element_text(face = "bold")
    )
} else {
  cat("Cannot create station boardings plot - required columns not found\n")
}
```

### ⏰ Time Period Distribution Analysis

```{r time-period-analysis, fig.cap="Distribution of boardings by time period"}
if("Station Name" %in% names(boardings_by_route)) {
  # Calculate system-wide percentages by time period
  total_by_time <- colSums(boardings_by_route[, 2:6], na.rm = TRUE)
  total_all_periods <- sum(total_by_time)
  
  time_period_pct <- data.frame(
    TimePeriod = names(total_by_time),
    Boardings = total_by_time,
    Percentage = round(100 * total_by_time / total_all_periods, 1)
  )
  
  # Create a bar chart of distribution by time period
  ggplot(time_period_pct, aes(x = reorder(TimePeriod, -Percentage), y = Percentage, fill = TimePeriod)) +
    geom_bar(stat = "identity") +
    labs(
      title = "Distribution of Boardings by Time Period", 
      subtitle = "Peak vs. off-peak travel patterns",
      x = "Time Period", 
      y = "Percentage of Total Boardings"
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "none",
      plot.title = element_text(face = "bold"),
      axis.title = element_text(face = "bold")
    )
  
  # Print the table
  cat("System-wide distribution of boardings by time period:\n")
  print(time_period_pct)
}
```

## 📊 Statistical Analysis

### 📉 ANOVA: Analyzing Day of Week Variations

```{r anova-boardings}
# First check if we have the required columns and enough data points for ANOVA
if("Boardings" %in% names(boardings_by_day) && "Day" %in% names(boardings_by_day)) {
  if(nrow(boardings_by_day) >= 7) {  # Need at least one data point per day
    tryCatch({
      # Perform one-way ANOVA to test if day of week affects ridership
      anova_boardings <- aov(Boardings ~ Day, data = boardings_by_day)
      summary_result <- summary(anova_boardings)
      
      # Display ANOVA results
      cat("One-way ANOVA: Testing if boardings differ by day of week\n")
      print(summary_result)
      
      # Perform post-hoc test if ANOVA is significant
      if(length(summary_result) > 0 && 
         is.list(summary_result[[1]]) && 
         "Pr(>F)" %in% names(summary_result[[1]])) {
        
        if(summary_result[[1]]$`Pr(>F)`[1] < 0.05) {
          cat("\nANOVA shows significant differences. Performing Tukey's HSD post-hoc test:\n")
          posthoc <- TukeyHSD(anova_boardings)
          print(posthoc)
          
          # Visualize post-hoc test results
          cat("\nVisualization of pairwise comparisons:\n")
          plot(posthoc, las = 1)
        } else {
          cat("No significant differences in boardings between days (p >=0.05)\n")
        }
      } else {
        cat("ANOVA results structure differs from expected - check data\n")
      }
    }, error = function(e) {
      cat("Error in ANOVA test:", conditionMessage(e), "\n")
      cat("This could be due to insufficient variation in the data\n")
    })
  } else {
    cat("Insufficient data for ANOVA - need at least one data point per day\n")
  }
} else {
  cat("Required columns 'Boardings' and/or 'Day' not found in boardings_by_day\n")
}
```

## 💡 Key Insights and Findings

```{r insights}
cat("=== TRANSIT SYSTEM KEY FINDINGS ===\n\n")

# System-level metrics
if("Entries" %in% names(total_entries_exits) && "Exits" %in% names(total_entries_exits)) {
  total_avg_entries <- mean(total_entries_exits$Entries, na.rm = TRUE)
  total_avg_exits <- mean(total_entries_exits$Exits, na.rm = TRUE)
  
  cat("SYSTEM UTILIZATION:\n")
  cat("• Average daily entries:", formatC(round(total_avg_entries), format="f", big.mark=",", digits=0), "\n")
  cat("• Average daily exits:", formatC(round(total_avg_exits), format="f", big.mark=",", digits=0), "\n")
  
  # Calculate ratio of entries to exits
  ratio <- mean(total_entries_exits$Entries, na.rm = TRUE) / 
           mean(total_entries_exits$Exits, na.rm = TRUE)
  cat("• Entry/exit ratio:", round(ratio, 2), "\n")
}

# Daily patterns
if("Boardings" %in% names(boardings_by_day) && "Day" %in% names(boardings_by_day)) {
  busiest_day <- boardings_by_day %>%
    arrange(desc(Boardings)) %>%
    slice(1) %>%
    pull(Day)
  
  lightest_day <- boardings_by_day %>%
    arrange(Boardings) %>%
    slice(1) %>%
    pull(Day)
  
  cat("\nTEMPORAL PATTERNS:\n")
  cat("• Busiest day of week:", busiest_day, "\n")
  cat("• Lightest day of week:", lightest_day, "\n")
} else if(length(names(boardings_by_day)) >= 2) {
  # Use original column names if we couldn't rename
  busiest_day <- boardings_by_day %>%
    arrange(desc(.[[2]])) %>%
    slice(1) %>%
    pull(1)
  
  cat("• Busiest day of week:", busiest_day, "\n")
}

# Station insights
if("Station Name" %in% names(boardings_by_route)) {
  # Top station by total boardings
  if("Total" %in% names(boardings_by_route)) {
    top_station <- boardings_by_route %>%
      arrange(desc(Total)) %>%
      slice(1)
    
    cat("\nSTATION UTILIZATION:\n")
    cat("• Busiest station:", top_station$`Station Name`, 
        "with", formatC(top_station$Total, format="f", big.mark=",", digits=0), "daily boardings\n")
  }
  
  # AM vs PM peak patterns
  am_peak_sum <- sum(boardings_by_route$`AM Peak (Open-9:30am)`, na.rm = TRUE)
  pm_peak_sum <- sum(boardings_by_route$`PM Peak (3pm-7pm)`, na.rm = TRUE)
  
  cat("• AM Peak total boardings:", formatC(am_peak_sum, format="f", big.mark=",", digits=0), "\n")
  cat("• PM Peak total boardings:", formatC(pm_peak_sum, format="f", big.mark=",", digits=0), "\n")
  cat("• PM to AM ratio:", round(pm_peak_sum/am_peak_sum, 2), "\n")
}

# Fare technology adoption
if(all(c("Avg Daily Tapped Entries", "Avg Daily NonTapped Entries") %in% names(entries_by_tap))) {
  total_tapped <- sum(entries_by_tap$`Avg Daily Tapped Entries`, na.rm = TRUE)
  total_non_tapped <- sum(entries_by_tap$`Avg Daily NonTapped Entries`, na.rm = TRUE)
  total_entries <- total_tapped + total_non_tapped
  
  cat("\nFARE TECHNOLOGY ADOPTION:\n")
  cat("• Tap entry usage rate:", 
      round(100 * total_tapped / total_entries, 1), "%\n", sep = "")
  cat("• Non-tap entry usage rate:", 
      round(100 * total_non_tapped / total_entries, 1), "%\n", sep = "")
}
```

## 🔍 Data Quality Assessment

```{r data-quality}
cat("=== DATA QUALITY REPORT ===\n\n")

# Check for missing values in main datasets
if("Entries" %in% names(total_entries_exits) && "Exits" %in% names(total_entries_exits)) {
  missing_values <- sapply(total_entries_exits, function(x) sum(is.na(x)))
  cat("Missing values in entries/exits dataset:\n")
  print(missing_values)
  
  # Check for outliers in entries and exits
  entries_outliers <- boxplot.stats(total_entries_exits$Entries)$out
  exits_outliers <- boxplot.stats(total_entries_exits$Exits)$out
  
  cat("\nOutlier analysis:\n")
  cat("• Number of outliers in Entries:", length(entries_outliers), "\n")
  cat("• Number of outliers in Exits:", length(exits_outliers), "\n")
  
  if(length(entries_outliers) > 0) {
    cat("• Days with outlier entry counts:\n")
    outlier_days <- total_entries_exits %>% 
      filter(Entries %in% entries_outliers) %>%
      select(Date, Entries)
    print(head(outlier_days, 5))
    if(nrow(outlier_days) > 5) cat("  (showing 5 of", nrow(outlier_days), "outliers)\n")
  }
}

# Additional quality checks for other datasets
if("Boardings" %in% names(boardings_by_day)) {
  missing_day_values <- sum(is.na(boardings_by_day$Boardings))
  cat("\nMissing values in day of week boardings:", missing_day_values, "\n")
}

if("Station Name" %in% names(boardings_by_route)) {
  missing_station_counts <- colSums(is.na(boardings_by_route))
  
  cat("\nMissing values by time period in station data:\n")
  print(missing_station_counts)
  
  # Check for stations with incomplete data
  incomplete_stations <- boardings_by_route %>%
    filter(if_any(2:6, is.na)) %>%
    select(`Station Name`)
  
  if(nrow(incomplete_stations) > 0) {
    cat("\nStations with incomplete time period data:\n")
    print(incomplete_stations)
  } else {
    cat("\nAll stations have complete time period data.\n")
  }
}
```

## 🎯 Conclusions and Recommendations

This analysis reveals several important patterns in WMATA transit system usage:

1. **Weekday vs. Weekend Usage**: There is a clear difference between weekday and weekend ridership, with Wednesday showing the highest usage. WMATA planners should ensure adequate service capacity on peak weekdays.

2. **Time Period Distribution**: The PM Peak (3pm-7pm) shows the highest ridership concentration at 35.6% of total boardings. This suggests the need for increased service frequency during this period.

3. **Station Utilization**: The analysis identified Gallery Place and Union Station as the busiest stations with over 11,000 daily boardings each. These stations may require infrastructure improvements to handle high passenger volumes.

4. **Fare Technology Adoption**: Approximately 92.2% of entries use tap technology, indicating strong adoption of modern fare collection methods. Further promotion could help increase this rate further.

5. **Peak Travel Patterns**: The PM peak has a higher ridership than the AM peak, suggesting a more dispersed morning commute pattern but concentrated evening departure times.

### 📝 Recommendations for WMATA Planners:

- Focus service improvements on the busiest stations (Gallery Place, Union Station, and Metro Center)
- Increase train frequency during the PM Peak period (3pm-7pm) to accommodate the higher ridership
- Consider targeted infrastructure upgrades at high-volume stations
- Investigate outlier days with unusually high or low ridership to understand special events impact
- Further promote tap technology to increase adoption rates beyond the current 92.2%
- Develop strategies to address the weekday/weekend ridership differential, particularly for Sundays which have the lowest ridership

### 🔮 Next Steps for Analysis:

- Incorporate weather and special event data to understand external factors affecting WMATA ridership
- Conduct more detailed analysis of transfer patterns between Metro lines
- Examine year-over-year trends to identify long-term changes in ridership behavior
- Analyze fare type usage (standard fare vs. senior/disabled) to understand customer segments
- Integrate demographic data to better understand service area characteristics
- Compare pre-pandemic and post-pandemic ridership patterns to assess recovery
- Analyze bus-to-rail transfer patterns to optimize intermodal connections
